---
title: "Getting Started with SArf"
author: "Kevin Credit"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with SArf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

This vignette demonstrates how to use the SArf package for spatial autoregressive random forest analysis from start to finish.

# Installation

## One-Time Setup

```{r install}
# Install development tools
install.packages("devtools")
install.packages("roxygen2")

# Install SArf dependencies
install.packages(c(
  "ranger", "sf", "spdep", "spatialreg", "blockCV", "ALEPlot",
  "ggplot2", "dplyr", "tidyr", "leaflet", "htmltools"
))

# Install SArf from GitHub
devtools::install_github("kcredit/SArf")

# Or from local directory
# setwd("~/path/to/SArf")
# devtools::document()
# devtools::install()

# IMPORTANT: Restart R session after installation!
```

# Load Libraries

```{r libraries}
library(SArf)
library(sf)
library(dplyr)

# Note: Only load these three to avoid conflicts!
```

# Real Example: Health Rating Index for Dublin

This example uses the full dataset from the Health Rating Index project.

**Project Background:** SArf was developed as part of research analyzing environmental health burdens and benefits across Dublin, Ireland. The full project is available at https://github.com/kcredit/health-rating-index

## Load Data

```{r dublin_example}
# Load included data
data_path <- system.file("extdata", "model_data.shp", package = "SArf")
data <- st_read(data_path)

# Check the data
cat("Dataset size:", nrow(data), "small areas\n")
summary(data$HRI_gaus_n)

# Variables included:
# - HRI_gaus_n: Health Rating Index (outcome)
# - In22_ED: Deprivation index
# - NoAuto_p: % households without car
# - POPD: Population density
# - log_dist: Log distance to city center
# - ov60: % over 60 years old
# - nonIrish: % non-Irish nationals

# Clean the data
data_clean <- data %>%
  # Remove missing outcome
  filter(!is.na(HRI_gaus_n)) %>%
  # Ensure numeric
  mutate(your_outcome_variable = as.numeric(HRI_gaus_n)) %>%
  # Remove missing predictors (add all your predictor names)
  filter(complete.cases(HRI_gaus_n, In22_ED, NoAuto_p, POPD, log_dist, ov60, nonIrish)) %>%
  # Fix invalid geometries
  filter(st_is_valid(.))

# Fix duplicate coordinates (important!)
st_geometry(data_clean) <- st_jitter(st_geometry(data_clean), amount = 0.00001)

# Transform to projected CRS (required for spatial blocking)
data_clean <- st_transform(data_clean, 3857)

# Check data
cat("Cleaned data:", nrow(data_clean), "rows\n")
summary(data_clean$HRI_gaus_p)

cat("Any NAs?", sum(is.na(data_clean$HRI_gaus_p)), "\n")
```

## Calculate Parameters

```{r calc_params}
n_rows <- nrow(data_clean)
cat("Dataset size:", n_rows, "rows\n")

# Choose parameters based on size
if (n_rows < 1000) {
  # Small dataset - full analysis
  n_folds <- 5
  n_bootstrap <- 20
  num_trees <- 500
  cat("Using full analysis parameters\n")
  
} else if (n_rows < 3000) {
  # Medium dataset - moderate
  n_folds <- 5
  n_bootstrap <- 10
  num_trees <- 500
  cat("Using moderate analysis parameters\n")
  
} else {
  # Very large - sample first
  cat("Very large dataset. Sampling 3000 rows...\n")
  set.seed(1111)
  data <- slice_sample(data_clean, n = 3000)
  n_folds <- 3
  n_bootstrap <- 5
  num_trees <- 250
}

cat("If your data are very large, you can alternatively (or additionally) increase the block range (rough guides):
    Neighbourhood: 500-1000m
    City: 5000-10000m
    County: 12500-25000m
    Region/Country: 50000-100000m", "\n")

# Get bounding box of your data
bbox <- st_bbox(data)

# Calculate width and height (in meters if projected)
width <- bbox["xmax"] - bbox["xmin"]
height <- bbox["ymax"] - bbox["ymin"]

# Use smaller dimension (more conservative)
min_dimension <- min(width, height)/3

```

## Run Analysis

```{r run_dublin}
# Run SArf analysis
results <- SArf(
  HRI_gaus_n ~ In22_ED + NoAuto_p + POPD + log_dist + ov60 + nonIrish,
  data = data,
  k_neighbors = 10,            # Number of neighbors for spatial weights
  n_folds = n_folds,           # Spatial CV folds
  n_bootstrap = n_bootstrap,   # Bootstrap iterations
  num_trees = num_trees,       # Trees in random forest
  create_map = TRUE,           # Spatial folds map
  block_range = min_dimension, # Size of the spatial blocks in units of metres
  verbose = TRUE               # Print progress
)
```

## View Results

```{r view_dublin_results}
# Print summary
print(results)

# Model comparison
results$model_comparison

# Variable importance
results$importance_plot

# ALE plots
results$ale_plots

# Interactive map
results$leaflet_map
```

## Save Outputs

```{r save_outputs}
# Create output directory
dir.create("output", showWarnings = FALSE)

# Save plots
ggsave("output/moran_plot.png", results$moran_plot, width = 6, height = 5)
ggsave("output/importance_plot.png", results$importance_plot, width = 8, height = 6)
ggsave("output/ale_plots.png", results$ale_plots, width = 10, height = 8)

# Save tables
write.csv(results$model_comparison, "output/model_comparison.csv", row.names = FALSE)
write.csv(results$variable_importance, "output/variable_importance.csv", row.names = FALSE)
write.csv(results$ale_results, "output/ale_data.csv", row.names = FALSE)

# Save map
# library(htmlwidgets)
# saveWidget(results$leaflet_map, "output/results_map.html")

cat("All outputs saved to output/ directory\n")
```


## Full Project

For the complete analysis and reproducible code:
- **GitHub:** https://github.com/kcredit/health-rating-index
- **Publication:** Credit et al. (2025), GISRUK Conference Proceedings
- **DOI:** 10.5281/zenodo.15183740

# Interpretation Guide

## Model Comparison

- **RMSE**: Lower is better
- **R²**: Higher is better (0-1 scale)
- **Moran's I**: Close to 0 indicates no spatial autocorrelation in residuals
- **Typically**: RF > spatial models for R², spatial models remove autocorrelation

## Variable Importance

- High mean = strong predictor
- Narrow CI = consistent across folds
- spatial_lag often important (neighborhood effects)

## ALE Plots

- Positive slope = positive effect
- Negative slope = negative effect
- Curves = non-linear relationships
- Confidence bands = uncertainty

# Tips and Best Practices

**Dataset Size:**
- < 1000: Full parameters
- 1000-3000: Moderate parameters
- 3000: Sample or reduce parameters

**Block Range:**
- Too small → "same area" error
- Too large → empty folds
- Start with min(width, height) / 3

**Performance:**
- 500 obs: ~5 min
- 1000 obs: ~10 min
- 2000 obs: ~20 min
- 3000 obs: ~45 min

**Avoid Conflicts:**
- Only load: SArf, sf, dplyr
- Don't load ranger, spdep, etc. directly

# Troubleshooting

**"All points from same area"**
- Increase block_range
- Reduce n_folds

**Moran's I shows NA**
- Check for missing values
- Ensure no duplicate coordinates

**Very slow**
- Sample to 2000-3000 rows
- Reduce n_bootstrap to 3-5

# Citation

```
Credit, K. (2025). SArf: Spatial Autoregressive 
Random Forest. https://github.com/YOUR-USERNAME/SArf

Credit, K., Kumar, D., & Eccles, E. (2025). Exploring the 
transport-health-environment nexus through a new 'Health Rating Index'. 
Proceedings of the 33rd GISRUK Conference. DOI: 10.5281/zenodo.15183740
```

# Session Info

```{r sessioninfo}
sessionInfo()
```
